The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-07-29 paper/paper.Rnw
Fixed math notation + added corrections from gosta dn bruno

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
.
% explain more in detail white income inequality is linked to to the achievement gap
% gosta page 3

@@

%

@@

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves. To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

% Talk in more detail about the ESCS index - is it a factor analysis? What's the methodology like?

@@

% We then use it to calculate the mean achievement and (standard error or standard deviation)?
For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps. The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization). We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..). "Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares. Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}. The result is a SES gap from an ordinal variable of interest.

@@

for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.

We then adjust \begin{math} \hat{Y_i} \end{math} by

% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
\begin{equation}
\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
\end{equation}

% Change the hat for something else because it makes the notation confusing.

@@

In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{Y_i} \end{math}) by

% As well, updated y hat inside the equation with the updated y from the
% equation above. And then change the hat from the left hand side with a new indicator.

\begin{equation}
\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
\end{equation}

>>>>

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. \citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children. This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.

However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
.

@@

@@

Yet one of the most relevant variables for our study is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}. This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.

The ESCS was derived from a principal component analysis of standardised variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status. All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only. However, for the purpose of reporting the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries \citep{pisa_2015_results}.

To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.

@@

For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps. The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization). We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..). "Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares. Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}. The result is a SES gap from an ordinal variable of interest.

@@

for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011}) weighted by the student sample weights.

We then adjust \begin{math} \hat{\gamma_i} \end{math} by

% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
\begin{equation}
\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
\end{equation}

@@

In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by

\begin{equation}
\hat{\gamma_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
\end{equation}
